Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-09 analysis/paper/paper.Rmd
Draft MS sent to co-authors

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for more years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
original approach, and @mcloughlin_declines_2003,
@hervieux_widespread_2013, @eacker_web-based_2019, @lamb_effectiveness_2024 and @dalgarno_bboutools_2024 for
examples of application to caribou). Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain
uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the
analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
population models, available demographic or population data (e.g.
abundance surveys, vital rate estimates, trend data) are combined into a
single analysis that can permit novel parameter estimation, and improve
precision, and account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. These models are proving broadly useful for
rapid detection of population declines, evaluation of management
actions, and viability analysis of critically endangered species
[@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models
informed only by local demographic data could be used to inflate
uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou. In our simple proof-of-concept model, information is 
shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. The model represents an important improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation
process.

@@

## Caribou population monitoring data and composition survey bias correction

@@

To derive a composition survey bias correction term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
uncollared adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-corrected recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias correction term $c$ that is integrated into
the Bayesian population model and simulations (below):
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

For this investigation we assume collars are deployed in January, and
last for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of uncollared to collared adults in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
[@johnson_science_2020] so we call this the "national" model. Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data with priors from the national model, and that in the absence of local data Bayesian model outcomes match expectations from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

### A model for simulating example population dynamics informed by national disturbance-demographic relationships

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio and misidentification biases
(Supplement B) to get maximum
potential recruitment of female calves per cow $\dot{X}_t$:

@@

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females
and the adjusted recruitment rate:

\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}).
  (\#eq:dotJ)
\end{equation}

@@

### Simulation of observed survival and recruitment

@@

Where appropriate, the Bayesian population model and national
demographic model share the same structure, and we have selected
notation to highlight the similarities. The difference between equations
\@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
simulate observations. In the Bayesian model, in contrast, observations
are provided, and used to estimate the unobserved recruitment rate
$R_t$.

@@

To account for sex ratio of calves and additional uncertainty caused by unknown
misidentification errors (Supplement B) the
estimated recruitment probability is adjusted to get expected female
recruits per female: 
\begin{equation}
X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias correction term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{ c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2.

Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1 [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression model estimates
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

As in the simulations, we assume a simple demographic model, with a
census in late winter of surviving females and new recruits to the
post-juvenile (aged 1+) class. Expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$. For reference, all variables and parameters are defined in supplemental Table A1.

Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\eta_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\eta_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\eta_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

@@

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), and we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are less
precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
left side of right panels in Figure \@ref(fig:examplePredictions)). At
lower levels of disturbance, local monitoring can reduce uncertainty,
there are diminishing returns as the number of collars deployed or
the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Adding unknown bias increases uncertainty, but these
increases are not large for the set of parameters we investigate
(Supplement F).

>>>>

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019, @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022, @eacker_web-based_2019, @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991;@hatter_revisiting_2020] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020; @dalgarno_bboutools_2024].
Various efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. An advantage of Bayesian demographic models is that they account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024] have not thus far included demographic-disturbance relationships, and we are concerned that uncertainty in models
informed only by local demographic data could be used to delay conservation action for poorly monitored populations. We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships [@johnson_science_2020] with
available local demographic data in a simple Bayesian population model
for boreal caribou.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012], none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb_effectiveness_2024; @dalgarno_bboutools_2024]. We address this gap with a bias term derived from a simple model of the recruitment survey observation
process.

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

## Caribou population monitoring data and composition survey errors

@@

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
uncollared adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-corrected recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

In simulated monitoring scenarios, we assume collars are deployed in January, and
last for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of uncollared to collared adults in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
[@johnson_science_2020] so we call this the "national" model. Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data, and that prior predictions meet expectations from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$.

Realized recruitment rate varies with population density
[@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
and the maximum
potential recruitment of female calves per cow ($\dot{X}_t$):

\begin{equation}
  \dot{J}_{t} \sim \text{Binomial}(\dot{W}_t,\dot{X}_t[p_0-(p_0-p_k)(\frac{\dot{W}_t}{k})^b]\frac{\dot{W}_t}{\dot{W}_t+a}),
  (\#eq:dotJ)
\end{equation}

The observable calf:cow
ratio $\dot{R}_t$ is adjusted for sex ratio and composition survey errors
(Supplement B) to get the maximum
potential recruitment of female calves per cow $\dot{X}_t$:

@@

@@

@@

Where appropriate, the Bayesian and national population models share the same structure, and we have selected notation to highlight the similarities. The difference between equations
\@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an
unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to
simulate observations. In the Bayesian model, in contrast, observations
are provided, and used to estimate the unobserved recruitment rate
$R_t$.

@@

To account for sex ratio of calves and additional uncertainty caused by composition survey errors errors (Supplement B) estimated recruitment is adjusted to get expected female
recruits per female: 
\begin{equation}
X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias term $c$ depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2.

Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
estimates [@kaplan_nonparametric_1958] of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

Again assuming a simple demographic model with a
census in late winter of surviving females and new recruits to the
post-juvenile class [@hatter_moose_1991; @hatter_revisiting_2020], the expected population growth rate (without interannual variation or demographic stochasticity) is $\tilde{\lambda}_t=\tilde{S}_t(1+c\tilde{R}_t/2)$.

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression models
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

Following Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\eta_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\eta_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\eta_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

@@

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), but we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are less
precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
left side of right panels in Figure \@ref(fig:examplePredictions)). At
lower levels of disturbance, local monitoring can reduce uncertainty, but
there are diminishing returns as the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Acknowledging the possibility of composition survey errors increases uncertainty, but these increases are not large for the set of parameters we investigate
(Supplement F).

```{r examplePredictions, fig.cap="Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation. In this example scenario anthropogenic disturbance increases by 2\\% per year from 0 to 100\\%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95\\% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5\\% and 97.5\\% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.", cache=TRUE}
